The Concurrent Calculi Formalisation Benchmark

### 2024-01-30 short-paper/short-paper.tex
small fixes, comments

\begin{abstract}
  The POPLMark challenge and its follow-up POPLMark Re\-loaded sparked a flurry of
  work on machine-checked proofs, and fostered the adoption of proof
  mechanisation in programming language research. However, both POPLMark and
  POPLMark Reloaded were purposely limited in scope, with benchmark problems
  that do not address concurrency-related issues.
  %
  For this reason, we propose a new collection of benchmark problems with a
  focus on the challenges and issues that typically arise when mechanising
  message-passing concurrency using process calculi.  Our benchmark problems
  address three key topics: linearity, scope extrusion, and coinductive
  reasoning.  The goal of this new challenge is to clarify, compare, and advance
  the state of the art, fostering the adoption of proof mechanisation in future
  research on message-passing concurrency.

@@

\section{Introduction}
By establishing a common ground for systematically evaluating mechanisation techniques, 
the POPLMark challenge~\cite{POPLMark} accelerated progress in 
the development of proof assistants, libraries, and best practices. In doing so, it 
spearheaded a shift in the programming language research community towards 
publications that include mechanised proofs. 
Its follow-up challenge, POPLMark Reloaded~\cite{POPLMarkReloaded}, 
facilitated the development of best practices and tool support for proofs 
using logical relations. These challenges were deliberately limited in scope, 
while their authors called for future investigations into such areas as 
coinduction and linear environments. In this spirit, we propose a 
new collection of benchmark problems specifically designed to address typical 
issues that arise when mechanising message-passing concurrency using process 
calculi.

@@

To this end, our benchmark problems (discussed in detail in \cref{app:challenges}) are designed to focus on three key aspects which typically cause difficulties 
when mechanising concurrency theory: \emph{linearity}, \emph{scope extrusion}, 
and \emph{coinductive reasoning} for infinite behaviours. These are certainly not the only factors that may cause issues, but they
are fundamental to concurrency theory, and emerge in most mechanisations.
The mechanisation of novel research results often requires addressing more than
just one of these aspects at once; we see the combination of techniques
as a next step beyond the scope of this challenge (we discuss this in
\cref{sec:going-beyond}).
% Our challenge problems are designed to stimulate progress on answering these questions by providing simple environments in which to demonstrate, evaluate, and compare the many tools and techniques available.
% While the challenge problems themselves are thus not of scientific interest, we expect that the best practices and mechanisations developed by solving the challenge problems will also be useful for future research.

Like the authors of the POPLMark challenges, we seek to
answer several questions:
\begin{enumerate}[label=\textbf{(Q\arabic*)},leftmargin=10mm]
\item\label{item:rq1} What is the current state of the art in mechanising process calculi?
\item\label{item:rq2} What techniques and best practices can be recommended when starting formalisation projects within concurrency theory?
\item\label{item:rq3} What improvements are needed to make mechanisation tools more user-friendly with regards to the issues faced when mechanising process calculi?
\end{enumerate}

@@

\paragraph{Outline of this paper.}
We discuss our design choices in \cref{sec:design-discussion}, and possible extensions and further topics in \cref{sec:going-beyond}.
The challenges themselves may be found in \cref{app:challenges}.

@@

\section{Design of the benchmark}\label{sec:design-discussion}

@@

\subsubsection{Evaluation criteria.}
The idea behind our benchmark is to obtain evidence towards answering questions \cref{item:rq1,item:rq2,item:rq3}. We are therefore interested in not only the solutions, but alsoin the experience of solving the challenges with the chosen approach.
Solutions to our challenges should be compared on three measures:
\begin{enumerate}
\item Mechanisation overhead: the amount of manually written infrastructure and setup needed to express the definitions in the mechanisation;
\item Adequacy of the formal statements in the mechanisation: whether the proven theorems are easily recognisable as the theorems from the challenge; and
\item Cost of entry for the tools and techniques employed: the difficulty of learning to use the techniques.
\end{enumerate}
Solutions to our challenges do not need to strictly follow the definitions and lemmata set out in the problem text, but solutions which deviate from the original challenges must present more elaborate argumentation for their adequacy.

@@

\vspace{-1mm}%
\paragraph{Extending and combining the current challenges.}
While the proposed challenges try to minimise the overlap of the techniques
and aspects they explore (linearity, scope extrusion, coinduction), most
research on concurrency theory combines some or all of these. A second round of formalisations might
entail combining the features of the challenges.
In \cref{tab:extensions} we consider the following three extensions:
\emph{(1)} the addition of choice to the type level and the process level,
\emph{(2)} the addition of recursion and (for typed systems) recursive types,
and \emph{(3)} the
addition of channel delegation to those calculi that do not already
support it.

@@

\vspace{-1mm}%
\paragraph{Addressing new calculi and new features.}
Extending in this direction
means proposing new challenges covering different features of
process calculi, or different proof mechanisation features.
Some interesting aspects of message-passing calculi to explore in
further challenges could be multiparty session types and
choreographies, as their meta-theory includes aspects
that we do not cover here (\eg: well-formed global types,
projection as a partial function). Also, one could easily
design a challenge that goes beyond our coinduction challenge, by
exploring different notions of bisimilarity (barbed, weak, \etc)
and trace equivalence.

>>>>

\begin{abstract}
  The POPLMark challenge and its follow-up POPLMark Re\-loaded sparked a flurry of
  work on machine-checked proofs, and fostered the adoption of proof
  mechanisation in programming language research. However, both POPLMark and
  POPLMark Reloaded were purposely limited in scope, with benchmark problems
  that do not address concurrency-related issues.
  %
  For this reason, we propose a new collection of benchmark problems with a
  focus on the challenges and issues that typically arise when mechanising
  message-passing concurrency using process calculi.\luca{wondering whether it would make sense to say ``\dots using formal models of concurrent and distributed programming languages in the form of process calculi'' or something similar to stress that process calculi are not the subject of the challenges, but ``only'' an appropriate tool for this kind of endeavor}  Our benchmark problems
  address three key topics: linearity, scope extrusion, and coinductive
  reasoning.  The goal of this new challenge is to clarify, compare, and advance
  the state of the art, fostering the adoption of proof mechanisation in future
  research on message-passing concurrency.

@@

\section{Introduction}
By establishing a common ground for systematically evaluating mechanisation techniques, 
the POPLMark challenge~\cite{POPLMark} accelerated progress in 
the development of proof assistants, libraries, and best practices. In doing so, it 
spearheaded a shift in the programming language research community towards 
publications that include mechanised proofs. 
Its follow-up challenge, POPLMark Reloaded~\cite{POPLMarkReloaded}, 
facilitated the development of best practices and tool support for proofs 
using logical relations. These challenges were deliberately limited in scope, 
while their authors called for future investigations into such areas as 
coinduction and linear environments. In this spirit, we propose a 
new collection of benchmark problems specifically designed to address typical 
issues that arise when mechanising message-passing concurrency using process 
calculi.\luca{see comment in abstract}

@@

To this end, our benchmark problems (discussed in detail in \cref{app:challenges}) are designed to focus on three key aspects which typically cause difficulties 
when mechanising concurrency theory: \emph{linearity}, \emph{scope extrusion}, 
and \emph{coinductive reasoning} for infinite behaviours. These are certainly not the only factors that may cause issues, but they
are fundamental to concurrency theory, and emerge in most mechanisations.
The mechanisation of novel research results often requires addressing more than
just one of these aspects at once; we see the combination of techniques
as a next step beyond the scope of this challenge (we discuss this further in
\Cref{sec:going-beyond}).
% Our challenge problems are designed to stimulate progress on answering these questions by providing simple environments in which to demonstrate, evaluate, and compare the many tools and techniques available.
% While the challenge problems themselves are thus not of scientific interest, we expect that the best practices and mechanisations developed by solving the challenge problems will also be useful for future research.

Like the authors of the POPLMark challenges, we seek to
answer several questions:
\begin{enumerate}[label=\textbf{(Q\arabic*)},leftmargin=10mm]
\item\label{item:rq1} What is the current state of the art in mechanising process calculi?
\item\label{item:rq2} Which techniques and best practices can be recommended when starting formalisation projects within concurrency theory?
\item\label{item:rq3} Which improvements are needed to make mechanisation tools more user-friendly with regards to the issues faced when mechanising process calculi?
\end{enumerate}

@@

\paragraph{Outline of this paper.}
We discuss our design choices in \Cref{sec:design-discussion}, and possible extensions and further topics in \Cref{sec:going-beyond}.
The challenges themselves may be found in \Cref{app:challenges}.

@@

\section{Design of the Benchmark}\label{sec:design-discussion}

@@

\subsubsection{Evaluation criteria.}
The idea behind our benchmark is to obtain evidence towards answering questions \cref{item:rq1,item:rq2,item:rq3}. We are therefore interested in not only the solutions, but also in the experience of solving the challenges with the chosen approach.
Solutions to our challenges should be compared on three measures:
\begin{enumerate}
\item Mechanisation overhead: the amount of manually written infrastructure and setup needed to express the definitions in the mechanisation;
\item Adequacy of the formal statements in the mechanisation: whether the proven theorems are easily recognisable as the theorems from the challenge; and
\item Cost of entry for the tools and techniques employed: the difficulty of learning to use the techniques.
\end{enumerate}
Solutions to our challenges do not need to strictly follow the definitions and lemmata set out in the problem text, but solutions which deviate from the original challenges must present more elaborate argumentation for their adequacy.

@@

\vspace{-1mm}%
\paragraph{Extending and combining the current challenges.}
While the proposed challenges try to minimise the overlap of the techniques
and aspects they explore (linearity, scope extrusion, coinduction), most
research on concurrency theory combines some or all of these. A second round of formalisations might
entail combining the features of the challenges.
In \Cref{tab:extensions} we consider the following three extensions:
\emph{(1)} the addition of choice to the type level and the process level,
\emph{(2)} the addition of recursion and (for typed systems) recursive types,
and \emph{(3)} the
addition of channel delegation to those calculi that do not already
support it.%
%
\luca{I would argue that support for recursive types and processes may also
impact the coinduction challenge. Indeed, some type systems with support for
recursion are more easily expressed \emph{coinductively}, by allowing possibly
infinite typing derivation. This application of coinduction is different from
that intended in the challenge, where it's focused on expressing notions of
process equivalence, and yet it is increasingly popular (see the related work on
infinitary linear logics) and occasionally seen with skepticism, which would
make a good case for formalization.}\luca{as a follow up to the previous
comment, if there is agreement that recursive types impact the coinduction
challenge then the relevance of Table 1 may be questionable because almost every
extension impacts almost every challenge. In case of need we may consider
removing the table in favor of more prose.}

@@

\vspace{-1mm}%
\paragraph{Addressing new calculi and new features.}
Extending in this direction
means proposing new challenges covering different features of
process calculi, or different proof mechanisation features.
Some interesting aspects of message-passing calculi to be explored in
further challenges could be multiparty session types and
choreographies, as their meta-theory includes aspects -- \eg well-formedness conditions on global types, partiality of end-point projection function, etc. --
that we do not cover here. Also, one could easily
design a challenge that goes beyond our coinduction challenge, by
exploring different notions of bisimilarity (barbed, weak, \etc)
and trace equivalence.
